---
title: "Exam 1 Group 2"
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.11.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---
##Section 1 - Brianna Duhart -Finished ##
# EX1 Group 2
#Note: Seperated the data into subsets for a collection of variables where job classification =1 and job classification =2. Then performed all of the analysis in lines 1-157 on both subsets to gain an understanding of the model with a focus on each subset. Use Wage_1 for job classification =1 dataset or Wage_2 for job classification =2 dataset. Completely finished with my section, please let me know if I should make any edits to my code or explainations. 

In this assignment we use the ch7-nonlin-lab as a template to further investigate the wage dataset from the ISLR2 library. The aim of this project is to re-analyze the wage data divided into two sub-data using jobclass as a classification variable

```{r chunk1}
library(ISLR2)
attach(Wage)
```

In this chunk we convert job class using numerical encoding into a clearly defined 1,2 classification variable. Where 1 is industrial and 2 is information
```{r chunk 1a}
head(Wage$jobclass)
class<- c(Wage$jobclass)
jobclass_factor <- factor(class, levels = c("1. Industrial", "2. Information"), labels = c(1, 2))
jobclass_num <- as.numeric(jobclass_factor)
head(jobclass_num)

```
After numerical encoding has been completed the original jobclass variable was replaced by the numerical version of jobclass to be used for the remainder of the project.
```{r chunk 1b}
Wage$jobclass<- jobclass_num
head(Wage$jobclass)
```
Now we will separate into Wage_1 and Wage_2 subsets of the data set to complete the project with this data set and understand the difference in modeling with different job classifications. 1 is Industrial and 2 is Information. 
```{r chunk 1c}
table(Wage$jobclass)
#1 industrial job classification
Wage_1 <- Wage[Wage$jobclass==1, ]
head(Wage_1)
#2 information job classification
Wage_2 <- Wage[Wage$jobclass==2, ]
head(Wage_2)
```

## Polynomial Regression and Step Functions
\def\caret{{^}}

We will create a linear model looking at age in the wage data separated using its two sub data sets with different classifications. We first fit this for a polynomial of 4.

```{r chunk2}
print("1st subset of Wage Data looking at Industrial Job class")
fit <- lm(wage ~ poly(age, 4), data = Wage_1)
coef(summary(fit))

#fit for information job classification
fit_ <- lm(wage ~ poly(age, 4), data = Wage_2)
print("2nd subset of Wage Data looking at Information Job class")
coef(summary(fit_))
```

The lm() function was used to predict the value of the outcome variable wage using a 4th degree polynomial in the age variable. Poly() was used to create a 4th degree polynomial which is returned as a matrix whose columns are a linear combination of the variables age, age^2, age^3, and age^4 using orthogonal values. This is important to condense the formula with powers of age. 


```{r chunk3}
fit2 <- lm(wage ~ poly(age, 4, raw = T), data = Wage_1)
coef(summary(fit2))

fit_2 <- lm(wage ~ poly(age, 4, raw = T), data = Wage_2)
coef(summary(fit_2))
```

This is another way to calculate the linear model fit using a 4th degree polynomial but instead of using the orthogonal set of basis functions the raw=T uses the raw/standard values. 


```{r chunk4}
fit2a <- lm(wage ~ age + I(age^2) + I(age^3) + I(age^4),
    data = Wage_1)
coef(fit2a)

fit_2a <- lm(wage ~ age + I(age^2) + I(age^3) + I(age^4),
    data = Wage_2)
coef(fit_2a)
```

This is another simple way to create a 4th degree polynomial by raising the chosen variable to the 4th, 3rd, 2nd, and 1st powers.

```{r chunk5}
fit2b <- lm(wage ~ cbind(age, age^2, age^3, age^4),
    data = Wage_1)

fit_2b <- lm(wage ~ cbind(age, age^2, age^3, age^4),
    data = Wage_2)
```


This is a more condensed approach using the cbind() function to build a matrix from the collection of variables. Then we create a grid of values for age in each subset and specify the standard errors desired. We separate the age and wage variables based on the different sub-data and predicted and fit the data to the model.

```{r chunk6}
age_1 <-Wage_1$age
age_2<-Wage_2$age
agelims_1 <- range(age_1)
age.grid_1 <- seq(from = agelims_1[1], to = agelims_1[2])
preds <- predict(fit, newdata = list(age = age.grid_1),
    se = TRUE)
se.bands1 <- cbind(preds$fit + 2 * preds$se.fit,
    preds$fit - 2 * preds$se.fit)


agelims_2 <- range(age_2)
age.grid_2 <- seq(from = agelims_2[1], to = agelims_2[2])
preds_2 <- predict(fit_, newdata = list(age= age.grid_2),
    se = TRUE)
se.bands_2 <- cbind(preds_2$fit + 2 * preds_2$se.fit,
    preds_2$fit - 2 * preds_2$se.fit)
```

Finally, we plot the data and add the fit from the degree-4 polynomial for job class =1 and job class =2 subset data. 

```{r chunk7}
wage_1 <- Wage_1$wage
wage_2 <- Wage_2$wage

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 1, 1),
    oma = c(0, 0, 4, 0))
plot(age_1, wage_1, xlim = agelims_1, cex = .5, col = "darkgrey")
title("JobClass = 1, Degree-4 Polynomial", outer = T)
lines(age.grid_1, preds$fit, lwd = 2, col = "blue")
matlines(age.grid_1, se.bands1, lwd = 1, col = "blue", lty = 3)


```

```{r chunk7b}
par(mfrow = c(1, 2), mar = c(4.5, 4.5, 1, 1),
    oma = c(0, 0, 4, 0))
plot(age_2, wage_2, xlim = agelims_2, cex = .5, col = "darkgrey")
title("JobClass = 2, Degree-4 Polynomial", outer = T)
lines(age.grid_2, preds_2$fit, lwd = 2, col = "blue")
matlines(age.grid_2, se.bands_2, lwd = 1, col = "blue", lty = 3)
```

Mar and oma are arguments to the par() function that enable changes to be made to the margins of the plots and the title function was used to create a title for each plot. We show that the fitted values obtained are identical within the same job classification group despite using an orthogonal set of basis functions and raw functions. 
```{r chunk8}
preds2_1 <- predict(fit2, newdata = list(age_1 = age.grid_1),
    se = TRUE)
max(abs(preds$fit - preds2_1$fit))

preds2_2 <- predict(fit_2, newdata = list(age_2 = age.grid_2),
    se = TRUE)
max(abs(preds_2$fit - preds2_2$fit))
```
We saw a very small difference 10^-11 between the two models using orthogonal versus raw basis functions within the same job classification group models. This suggests that there is a negligible difference between the two choices so the fitted values can be considered identical.

## ANOVA Section - Yuelin - Finished ##
In order to choose the appropriate polynomial degree, we will compare models via hypothesis tests, starting from a linear model, then increasing model complexity to degree-5 polynomials, to decide the simplest model to explain the relationship between `wage` and `age` for each job class. 
We use the `anova()` function to get the value of the F-test in \define{analysis of variance} (ANOVA), to compare nested polynomial models and see whether those higher-degree terms improve model fit by reducing prediction error (RSS) significantly. If the p-value for the null hypothesis is smaller than our desired significance level, adding polynomial terms improves the model performance. Otherwise, a larger p-value tells us that we fail to reject the null hypothesis, and the simpler model is sufficient. This allows us to select the simplest adequate model without overfitting. To perform the `anova()` function, we need to make sure these models are *nested* - the predictors in the simpler model $\mathcal{M}_1$ must be a subset of the predictors in the complex model $\mathcal{M}_2$. Here, we fit five different models to compare the p-values of the simpler model to the complex model by sequence.

```{r chunk9}
#For the Industrial group (JobClass = 1):
fit1_1 <- lm(wage ~ age, data = Wage_1)
fit2_1 <- lm(wage ~ poly(age, 2), data = Wage_1)
fit3_1 <- lm(wage ~ poly(age, 3), data = Wage_1)
fit4_1 <- lm(wage ~ poly(age, 4), data = Wage_1)
fit5_1 <- lm(wage ~ poly(age, 5), data = Wage_1)
anova(fit1_1, fit2_1, fit3_1, fit4_1, fit5_1)

#For the Information group (JobClass = 2):
fit1_2 <- lm(wage ~ age, data = Wage_2)
fit2_2 <- lm(wage ~ poly(age, 2), data = Wage_2)
fit3_2 <- lm(wage ~ poly(age, 3), data = Wage_2)
fit4_2 <- lm(wage ~ poly(age, 4), data = Wage_2)
fit5_2 <- lm(wage ~ poly(age, 5), data = Wage_2)
anova(fit1_2, fit2_2, fit3_2, fit4_2, fit5_2)
```


For the $Industrial group (JobClass = 1)$, the p-value comparing the linear model (Model 1) to the quadratic model (Model 2) is extremely small ($<$$2e-16$), indicating that a linear model is not sufficient. The p-values for adding higher polynomial terms (Model 3-5) are all greater than $0.05$, suggesting that additional polynomial terms of degree more than 2 do not improve the model fit significantly. Thus, a $quadratic (degree 2) model is sufficient for this $Industrial group$.
For the $Information group (JobClass = 2)$, both the quadratic term ($1.272e-12$) and the cubic term ($0.0039$) improve the model significantly. This indicates that neither a linear nor a quadratic model is sufficient. The p-values for the higher polynomial terms (degree 4 and 5) are greater than $0.05$, suggesting they do not improve the model fit. Thus, a $cubic (degree 3) model is sufficient$ for this $Information group$.
Different job classes show different levels of nonlinearity in the wage-age relationship, which requires different polynomial degrees for adequate modeling.

The `poly()` function can create orthogonal polynomials; we can obtain the p-values fast by using it instead of the `anova()` function.

```{r chunk10}
#For the Industrial group (JobClass = 1):
coef(summary(fit5_1))

#For the Information group (JobClass = 2):
coef(summary(fit5_2))
```

We can see that the p-values are the same, and the square of the  $t$-statistics are the same as the F-statistics from the `anova()` function; for example:

```{r chunk11}
(-2.891646)^2
```

Except for whether or not we used orthogonal polynomials, ANOVA also works when we have other predictors in the model as well. This allows us to evaluate whether high-order polynomial terms of $age$, in our case, still improve the model after adding other variables. Here is the example:

```{r chunk12}
#For the Industrial group (JobClass = 1):
fit.1_1 <- lm(wage ~ education + age, data = Wage_1)
fit.2_1 <- lm(wage ~ education + poly(age, 2), data = Wage_1)
fit.3_1 <- lm(wage ~ education + poly(age, 3), data = Wage_1)
anova(fit.1_1, fit.2_1, fit.3_1)

#For the Information group (JobClass = 2):
fit.1_2 <- lm(wage ~ education + age, data = Wage_2)
fit.2_2 <- lm(wage ~ education + poly(age, 2), data = Wage_2)
fit.3_2 <- lm(wage ~ education + poly(age, 3), data = Wage_2)
anova(fit.1_2, fit.2_2, fit.3_2)
```

This analysis allows us to determine whether adding quadratic or cubic terms of `age` provides a significant improvement in model fit after controlling for `education`.
For the $Industrial group$, higher-order terms more than degree 2 are not significant, indicating that a quadratic model remains sufficient.
For the $Information group$, the cubic term still provides a significant improvement, suggesting a degree 3 polynomial is sufficient.
Overall, this demonstrates that the conclusion about the polynomial degree remains stable even when adding more predictors to the model.

Next we consider predicting whether an individual earns more than $$250{,}000$ per year. We follow a similar approach as before, but define a binary response variable using `I(wage > 250)` and fit polynomial logistic regression models for each job class. Based on the ANOVA results above, we use a degree-2 polynomial for the Industrial group and a degree-3 polynomial for the Information group.

```{r chunk13}
#For the Industrial group (JobClass = 1):
fit_logit_1 <- glm(I(wage > 250) ~ poly(age, 2),
                   data = Wage_1, family = binomial)

#For the Information group (JobClass = 2):
fit_logit_2 <- glm(I(wage > 250) ~ poly(age, 3),
                   data = Wage_2, family = binomial)
```

The `I(wage > 250)` creates a binary response variable. Values greater than 250 are coded as 1 for `TRUE` and others as 0 for `FALSE`. This setting will fit the `glm()` function.

The default prediction type in `glm()` is `type = "link"`. To obtain the confidence intervals for $\Pr(Y=1|X)$, we transform the log odds scale we got from the *logit* 
\[
\log\left(\frac{\Pr(Y=1|X)}{1-\Pr(Y=1|X)}\right)=X\beta,
\]
 and the predictions given are of the form $X\hat\beta$.  The  standard errors given are also for $X \hat\beta$. In order to obtain confidence intervals for $\Pr(Y=1|X)$, we use the transformation
\[
\Pr(Y=1|X)=\frac{\exp(X\beta)}{1+\exp(X\beta)}.
\]

```{r chunk15}
#For the Industrial group (JobClass = 1):
preds_1 <- predict(fit_logit_1, newdata = list(age = age.grid_1), se = TRUE)
pfit_1 <- exp(preds_1$fit) / (1 + exp(preds_1$fit))
se.bands.logit_1 <- cbind(preds_1$fit + 2 * preds_1$se.fit,
                          preds_1$fit - 2 * preds_1$se.fit)
prob.bands_1 <- exp(se.bands.logit_1) / (1 + exp(se.bands.logit_1))

#For the Information group (JobClass = 2):
preds_2 <- predict(fit_logit_2, newdata = list(age = age.grid_2), se = TRUE)
pfit_2 <- exp(preds_2$fit) / (1 + exp(preds_2$fit))
se.bands.logit_2 <- cbind(preds_2$fit + 2 * preds_2$se.fit,
                          preds_2$fit - 2 * preds_2$se.fit)
prob.bands_2 <- exp(se.bands.logit_2) / (1 + exp(se.bands.logit_2))
```

Although we could compute the predicted probabilities directly by selecting the `type = "response"` in the `predict()` function, we do not use it when obtaining confidence intervals.

```{r chunk16}
#For the Industrial group (JobClass = 1):
preds_1 <- predict(fit_logit_1, newdata = list(age = age.grid_1),
                   type = "response", se = TRUE)

#For the Information group (JobClass = 2):
preds_2 <- predict(fit_logit_2, newdata = list(age = age.grid_2),
                   type = "response", se = TRUE)
```

Because the standard errors returned by `glm()` are calculated on the logit (link) scale, not on the probability scale. Directly computing the confidence intervals using these values after transformation will lead to invalid results, such as probabilities outside [0, 1] range. So, we compute confidence intervals on the logit scale, then transform them to probabilities to ensure all predicted values and intervals remain in the valid range.

Finally, the right-hand plots were made as follows:

```{r chunk17}
par(mfrow = c(1, 2))

#For the Industrial group (JobClass = 1):
plot(age_1, I(wage_1 > 250), xlim = agelims_1, type = "n",
     ylim = c(0, .2), main = "Industrial",
     xlab = "age", ylab = "Pr(wage > 250)")
points(jitter(age_1), I((wage_1 > 250)/5),
       cex = .5, pch = "|", col = "darkgrey")
lines(age.grid_1, pfit_1, col = "blue", lwd = 2)
matlines(age.grid_1, prob.bands_1, col = "blue", lty = 3)

#For the Information group (JobClass = 2):
plot(age_2, I(wage_2 > 250), xlim = agelims_2, type = "n",
     ylim = c(0, .2), main = "Information",
     xlab = "age", ylab = "Pr(wage > 250)")
points(jitter(age_2), I((wage_2 > 250)/5),
       cex = .5, pch = "|", col = "darkgrey")
lines(age.grid_2, pfit_2, col = "blue", lwd = 2)
matlines(age.grid_2, prob.bands_2, col = "blue", lty = 3)
```

In the plots, the gray points on the top are those `wage` more than $250$, those `wage` less than $250$ are the gray points on the bottom. The `jitter()` function is used here to jitter the `age` values to distinct observations with the same `age` value.

For the  $Industrial group$, the probability remains low across all ages and shows only a mild nonlinear pattern. There is a slight increase around middle age, followed by a decline, indicating that high wages are relatively rare in this group and not that strongly dependent on age. This pattern is consistent with the quadratic model selected earlier.
For the $Information group$, the probability of earning a high wage increases with age, then declines after middle age. Compared with the Industrial group, the overall probability is higher and shows a clearer nonlinear pattern. This demonstrates the use of the cubic model earlier.

Although the Industrial group and information group are modeled by the quadratic and cubic polynomial, now, as an alternative nonlinear modeling approach, we fit $step functions$ for each job class in order to compare a piecewise-constant fit with the smoother polynomial models. The same number of intervals is used for both groups for comparability.

To fit a step function, we divide $age$ into intervals using the `cut()` function, then fit separate models for each job class.

```{r chunk18}
#For the Industrial group (JobClass = 1):
table(cut(age_1, 4))
fit_step_1 <- lm(wage ~ cut(age, 4), data = Wage_1)
coef(summary(fit_step_1))

#For the Information group (JobClass = 2):
table(cut(age_2, 4))
fit_step_2 <- lm(wage ~ cut(age, 4), data = Wage_2)
coef(summary(fit_step_2))
```

The `cut()` function automatically selects cutoff points that divide the data into 4 groups. It picked the cutoff points at $33.5$, $49$, $64.5$, and $80.1$~years of age. We could also create our own cutoff points directly using the `breaks` option.
The function `cut()` orders a categorical variable; then the `lm()` function creates a set of dummy variables in the regression. The `< 33.5` in the age category is left out. 
For $Industrial group$, the average salary for those nder $33.5 years of age is the intercept coefficient of $$87{,}458$, and the other coefficients are the additional salary for those in the other age groups, in average. The result of p-values indicate that the second and third age groups (33.5-49 and 49-64.5) have significant average wages, compared to the baseline age group (<=33.5), the oldest age group (64.5-80.1) does not have significant difference from the baseline.
For the $Information group$, similarly, the two middle age groups (33.5-49 and 49-64.5) have significant higher wages compared to the baseline age group, in average. The oldest age group does not have significant difference from the baseline.
Overall, from both job classes, wages tend to reach to a peak at the middle age groups and then become variable at older ages. The significance in the oldest group also suggests higher uncertainty in that age range.

Comparing to polynomial models, step functions provide a simpler, piecewise approximation of the age and wage relationship. They can show general trends but lack smoothness and may not represent the underlying pattern, especially for the Information group where the relationship is more complex.



Now, we use the `splines` library to fit regression splines to obtain a smoother fit.
We use the `bs()` function to generate the entire matrix of basis functions to fit regression splines with the specified set of knots. The default cubic splines is using to fit `wage` to `age` for each job class:

```{r chunk19}
library(splines)
par(mfrow = c(1, 2))

#For the Industrial group (JobClass = 1):
fit_bs_1 <- lm(wage ~ bs(age, knots = c(25, 40, 60)), data = Wage_1)
pred_bs_1 <- predict(fit_bs_1, newdata = list(age = age.grid_1), se = TRUE)
plot(age_1, wage_1, col = "gray", main = "Industrial",
     xlab = "age", ylab = "wage")
lines(age.grid_1, pred_bs_1$fit, lwd = 2)
lines(age.grid_1, pred_bs_1$fit + 2 * pred_bs_1$se, lty = "dashed")
lines(age.grid_1, pred_bs_1$fit - 2 * pred_bs_1$se, lty = "dashed")

#For the Information group (JobClass = 2):
fit_bs_2 <- lm(wage ~ bs(age, knots = c(25, 40, 60)), data = Wage_2)
pred_bs_2 <- predict(fit_bs_2, newdata = list(age = age.grid_2), se = TRUE)
plot(age_2, wage_2, col = "gray", main = "Information",
     xlab = "age", ylab = "wage")
lines(age.grid_2, pred_bs_2$fit, lwd = 2)
lines(age.grid_2, pred_bs_2$fit + 2 * pred_bs_2$se, lty = "dashed")
lines(age.grid_2, pred_bs_2$fit - 2 * pred_bs_2$se, lty = "dashed")
```

We set up the knots at ages $25$, $40$, and $60$ for both job classes, in order to follow the original lab and ensure that spline models for the two groups are comparable directly. This produces a spline with six basis functions. 
We can also use the `df` option to create a spline with three knots at the same quantiles of the data. Because a cubic spline starts with four degrees of freedom, and each knot adds one more degree of freedom, a cubic spline with three knots has 7 degrees of freedom. Here, when we use `df = 6`, R selects the knot locations automatically to achieve that level of flexibility.

```{r chunk20}
#For the Industrial group (JobClass = 1):
dim(bs(age_1, knots = c(25, 40, 60)))
dim(bs(age_1, df = 6))
attr(bs(age_1, df = 6), "knots")

#For the Information group (JobClass = 2):
dim(bs(age_2, knots = c(25, 40, 60)))
dim(bs(age_2, df = 6))
attr(bs(age_2, df = 6), "knots")
```

For the $Industrial group$, `R` chooses knots at ages $32, 41$, and $50$ at 25th, 50th, and 75th percentiles of `age`. And for the $Information group$, it chooses knots at ages $35, 43$, and $51$ at the same percentiles of age. The differences indicate there is variation in the age groups between the two job classes. Same as the previous, the results show that the $Industrial group$ has a smoother and simpler pattern, while the $Information group$ has a more curved pattern. This also indicates that there is a stronger nonlinear relationship between age and wage.
The `bs()` function has a `degree` argument also, so we can fit splines of any degree to control flexibility.

We use the `ns()` function to fit a natural spline. 
For each job class, we fit a natural spline with four degrees of freedom.

## Section 3- Srinivasa Reddy ##
```{r chunk21}
fit2 <- lm(wage ~ ns(age, df = 4), data = Wage)
pred2 <- predict(fit2, newdata = list(age = age.grid),
     se = T)
plot(age, wage, col = "gray")
lines(age.grid, pred2$fit, col = "red", lwd = 2)
```

As with the `bs()` function, we could instead specify the knots directly using the `knots` option.

  In order to fit a smoothing spline, we use
the `smooth.spline()` function.
Figure 7.8  was produced with the following
code:

```{r chunk22}
plot(age, wage, xlim = agelims, cex = .5, col = "darkgrey")
title("Smoothing Spline")
fit <- smooth.spline(age, wage, df = 16)
fit2 <- smooth.spline(age, wage, cv = TRUE)
fit2$df
lines(fit, col = "red", lwd = 2)
lines(fit2, col = "blue", lwd = 2)
legend("topright", legend = c("16 DF", "6.8 DF"),
    col = c("red", "blue"), lty = 1, lwd = 2, cex = .8)
```

Notice that in the first call to `smooth.spline()`, we specified `df = 16`. The function then  determines which value of $\lambda$  leads to $16$ degrees of freedom.
 In the second call to `smooth.spline()`, we select the smoothness level by cross-validation;
 this results in a value of $\lambda$ that yields 6.8 degrees of freedom.

In order to perform local regression, we use the `loess()` function. 

```{r chunk23}
plot(age, wage, xlim = agelims, cex = .5, col = "darkgrey")
title("Local Regression")
fit <- loess(wage ~ age, span = .2, data = Wage)
fit2 <- loess(wage ~ age, span = .5, data = Wage)
lines(age.grid, predict(fit, data.frame(age = age.grid)),
    col = "red", lwd = 2)
lines(age.grid, predict(fit2, data.frame(age = age.grid)),
    col = "blue", lwd = 2)
legend("topright", legend = c("Span = 0.2", "Span = 0.5"),
    col = c("red", "blue"), lty = 1, lwd = 2, cex = .8)
```

Here we have performed local linear regression using spans of $0.2$ and $0.5$: that is, each neighborhood consists of 20\,\% or 50\,\% of the observations. The larger the span, the smoother the fit.
The `locfit` library can also be used for fitting local regression models in `R`.


## GAMs


We now fit a GAM to predict `wage` using natural spline functions of `lyear` and `age`,  treating `education` as a qualitative predictor, as in (7.16). Since this is just a big linear regression model using an appropriate choice of basis functions, we can simply do this using the `lm()` function.

```{r chunk24}
gam1 <- lm(wage ~ ns(year, 4) + ns(age, 5) + education,
    data = Wage)
```


We now fit the model (7.16) using smoothing splines rather than natural splines.  In order to fit more general sorts of GAMs, using smoothing splines or other components that cannot be expressed in terms of basis functions and then fit using least squares regression, we will need to use the `gam` library in `R`.

The `s()` function, which is part of the `gam` library, is used to indicate that we would like to use a smoothing spline. We specify that the function of `lyear` should have $4$ degrees of freedom, and that the function of `age` will have $5$ degrees of freedom. Since `education` is qualitative, we leave it as is, and it is converted into four dummy variables. We use the `gam()` function in order to fit a GAM using these components.
 All of the terms in (7.16) are fit simultaneously, taking each other into account to explain the response.

```{r chunk25}
library(gam)
gam.m3 <- gam(wage ~ s(year, 4) + s(age, 5) + education,
    data = Wage)
```


In order to produce  Figure 7.12, we simply call the `plot()` function:

```{r chunk26}
par(mfrow = c(1, 3))
plot(gam.m3, se = TRUE, col = "blue")
```

The generic `plot()` function recognizes that `gam.m3` is an
object of class `Gam`, and invokes the appropriate
`plot.Gam()` method.  Conveniently, even though
  `gam1` is not of class `Gam` but rather of class `lm`, we can {\em
  still} use `plot.Gam()` on it.  Figure 7.11 was produced using the following expression:

```{r chunk27}
plot.Gam(gam1, se = TRUE, col = "red")
```

Notice here we had to use `plot.Gam()` rather than the *generic* `plot()` function.

In these plots, the function of `lyear` looks rather linear. We can perform a series of ANOVA tests in order to determine which of these three models is best:  a GAM that excludes `lyear` ($\mathcal{M}_1$),  a GAM that uses a linear function of `lyear` ($\mathcal{M}_2$), or  a GAM that uses a spline function
of `lyear` ($\mathcal{M}_3$).

```{r chunk28}
gam.m1 <- gam(wage ~ s(age, 5) + education, data = Wage)
gam.m2 <- gam(wage ~ year + s(age, 5) + education,
    data = Wage)
anova(gam.m1, gam.m2, gam.m3, test = "F")
```

We find that there is compelling evidence that a GAM with a linear function of `lyear` is better than a GAM that does not include `lyear` at all \hbox{(p-value\,=\,0.00014).} However, there is no evidence that a non-linear function of `lyear` is needed (p-value\,=\,0.349).
In other words, based on the results of this ANOVA, $\mathcal{M}_2$ is preferred.

The `summary()` function produces a summary of the gam fit.

```{r chunk29}
summary(gam.m3)
```

##Section 4 -Basava Divya ##
The ``Anova for Parametric Effects" p-values clearly demonstrate that `year`, `age`, and `education` are all highly statistically significant, even when only assuming a linear relationship. Alternatively, the ``Anova for Nonparametric Effects" p-values for `year` and `age` correspond to a null hypothesis of a linear relationship versus the alternative of a non-linear relationship. The large p-value for `year` reinforces our conclusion from the ANOVA test that a linear function is adequate for this term. However, there is very clear evidence that a non-linear term is required for `age`.

We can make predictions
using the `predict()` method for the class `Gam`.
 Here we make predictions on the training set.

```{r chunk30}
preds <- predict(gam.m2, newdata = Wage)
```


We can also use local regression fits as building blocks in a GAM, using the `lo()` function.

```{r chunk31}
gam.lo <- gam(
    wage ~ s(year, df = 4) + lo(age, span = 0.7) + education,
    data = Wage
  )
plot.Gam(gam.lo, se = TRUE, col = "green")
```

Here we have used local regression for the `age` term, with a span of
$0.7$.
We can also use the `lo()` function to create interactions before calling the `gam()` function. For example,

```{r chunk32}
gam.lo.i <- gam(wage ~ lo(year, age, span = 0.5) + education,
    data = Wage)
```

fits a two-term model, in which the first term is an interaction between `lyear` and `age`, fit by a local regression surface. We can plot the resulting two-dimensional surface if we first install the `akima` package.

```{r chunk33}
library(akima)
plot(gam.lo.i)
```




In order to fit a logistic regression GAM, we once again use the `I()` function in constructing the binary response variable, and set `family=binomial`.

```{r chunk34}
gam.lr <- gam(
    I(wage > 250) ~ year + s(age, df = 5) + education,
    family = binomial, data = Wage
  )
par(mfrow = c(1, 3))
plot(gam.lr, se = T, col = "green")
```

It is easy to see that there are no high earners in the `< HS` category:

```{r chunk35}
table(education, I(wage > 250))
```

Hence, we fit a logistic regression GAM using all but this category. This provides more sensible results.

```{r chunk36}
gam.lr.s <- gam(
    I(wage > 250) ~ year + s(age, df = 5) + education,
    family = binomial, data = Wage,
    subset = (education != "1. < HS Grad")
  )
plot(gam.lr.s, se = T, col = "green")
```



